Cloudbreak: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\ Supplementary Material

### 2014-01-24 manuscript/bioinfo/document.tex
remove most todos

In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data. In particular, Google's MapReduce~\citep{Dean:2008p277} was designed to manage the storage and efficient processing of big data across clusters of commodity servers. Hadoop is an open source project of the Apache Foundation that provides an implementation of MapReduce. MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention. The Hadoop framework has been successfully applied to sequencing-related tasks including short read mapping~\citep{Schatz:2009p278} and SNP calling~\citep{Langmead:2009p1225}\todo{, and RNA-seq differential expression analysis~\citep{Langmead:2010p1268}}.

@@

The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments. \todo{This} We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair. The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.

@@

Choosing the correct threshold to set on the output of an SV calling algorithm can be difficult. The use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying thresholds. First, we characterized the predictions made by each algorithm at the threshold that gives them maximum sensitivity. \todo{For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low.} The results are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}. MoDIL and Cloudbreak exhibited the greatest recall for deletions. Cloudbreak also has high precision at this threshold, and discovers many small variants. For insertions, Cloudbreak has the highest recall, although recall is low for all four approaches. Cloudbreak again identifies many small variants. Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by RP mapping. We also used the ROC curves to characterize algorithm performance when a low false discovery rate is required. Supplementary Table 2\todo{\ref{Schr2DeletionPredsFDR10}} shows the total number of deletions found when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve. At this more stringent cutoff, Cloudbreak identifies more deletions in every size category than any other tool. Insertion performance never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table.\todo{ We also examined Cloudbreak's ability to detect events in repetitive regions of the genome, and found that it was similar to the other methods tested (Supplementary Tables \ref{SdeletionRepmaskpreds} and \ref{SinsertionRepmaskpreds}).}

@@

On simulated and real data sets, our approach exhibits high accuracy when compared to popular SV detection algorithms that run on traditional clusters and servers. Detection of insertions and deletions is an important area of research; \cite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans. \todo{More SV annotations are also required to assess the impact of SVs on non-coding elements~\citep{Mu:2011br}.}

>>>>

In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data. In particular, Google's MapReduce~\citep{Dean:2008p277} was designed to manage the storage and efficient processing of big data across clusters of commodity servers. Hadoop is an open source project of the Apache Foundation that provides an implementation of MapReduce. MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention. The Hadoop framework has been successfully applied to sequencing-related tasks including short read mapping~\citep{Schatz:2009p278} and SNP calling~\citep{Langmead:2009p1225}.

@@

The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments. We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair. The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.

@@

Choosing the correct threshold to set on the output of an SV calling algorithm can be difficult. The use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying thresholds. First, we characterized the predictions made by each algorithm at the threshold that gives them maximum sensitivity. The results are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}. MoDIL and Cloudbreak exhibited the greatest recall for deletions. Cloudbreak also has high precision at this threshold, and discovers many small variants. For insertions, Cloudbreak has the highest recall, although recall is low for all four approaches. Cloudbreak again identifies many small variants. Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by RP mapping. We also used the ROC curves to characterize algorithm performance when a low false discovery rate is required. Supplementary Table 2\todo{\ref{Schr2DeletionPredsFDR10}} shows the total number of deletions found when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve. At this more stringent cutoff, Cloudbreak identifies more deletions in every size category than any other tool. Insertion performance never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table.

@@

On simulated and real data sets, our approach exhibits high accuracy when compared to popular SV detection algorithms that run on traditional clusters and servers. Detection of insertions and deletions is an important area of research; \cite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.

### 2014-01-24 manuscript/bioinfo/supplementary.tex
additional rewrites
remove most todos

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 120 & 327 & 562 & 8059 \\ 
  \hline
  Cloudbreak  & \textbf{37} (\textbf{10}) & \textbf{180} (\textbf{25}) & \textbf{300} (\textbf{85}) & \textbf{1166} (\textbf{270}) \\ 
  Breakdancer & 29 (7) & 142 (3) & 192 (11) & 868 (21) \\
  GASVPro     & 16 (1) & 79 (1) & 79 (7) & 330 (16) \\
  DELLY-RP       & 21 (1) & 117 (1) & 152 (6) & 712 (5) \\
  DELLY-SR       & 0 (0) & 42 (0) & 27 (0) & 391 (0) \\
  Pindel      & 18 (9) & 112 (\textbf{25}) & 109 (2) & 536 (7) \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element. For both data sets we used the thresholds derived from finding the 10\% FDR level in the simulated data set. The proportion of deletions that overlap repetitive elements discovered by Cloudbreak is similar to that of the other methods.}
\label{deletionRepmaskpreds}
\end{table}

\begin{table}[h]
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 133 & 270 & 341 & 355 \\ 
  \hline
  Cloudbreak  & \textbf{32} (11) & \textbf{91} (\textbf{47}) & \textbf{169} (\textbf{61}) & \textbf{148} (\textbf{68}) \\ 
  Breakdancer & 17 (5) & 22 (6) & 82 (7) & 44 (9) \\
  Pindel      & 25 (\textbf{16}) & 54 (30) & 84 (35) & 82 (24) \\ 
  MoDIL      & 5 (0) & 16 (3) & NA & NA \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element. The maximum sensitivity cutoffs were used for both data sets. The proportion of insertions in repetitive elements discovered by Cloudbreak is similar to that of the other methods.}
\label{insertionRepmaskpreds}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{r|rrrrr}
 \hline
 Window Size & Calls & True Positives & Precision & Recall & F1 \\ 
 \hline
   1 & 274 & 228 & 0.832 & 0.57 & 0.677 \\ 
   5 & 268 & 228 & 0.851 & 0.57 & 0.683 \\ 
   10 & 258 & 223 & 0.864 & 0.557 & 0.678 \\ 
   25 & 240 & 217 & 0.904 & 0.542 & 0.678 \\ 
   50 & 215 & 194 & 0.902 & 0.485 & 0.631 \\ 
   100 & 162 & 141 & 0.87 & 0.352 & 0.502 \\  
\end{tabular}
\end{center}
\caption[Cloudbreak accuracy with varying window sizes.]{Accuracy measures for Cloudbreak on the Chromosome 2 simulated data set using different choices of window size. For each window size the same Cloudbreak score threshold was used (1.98).}
\label{chr2AccuracyByWindowSize}
\end{table}

>>>>

\newcommand{\todo}[1]{}

@@

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{r|rrrrr}
 \hline
 Window Size & Calls & True Positives & Precision & Recall & F1 \\ 
 \hline
   1 & 274 & 228 & 0.832 & 0.57 & 0.677 \\ 
   5 & 268 & 228 & 0.851 & 0.57 & 0.683 \\ 
   10 & 258 & 223 & 0.864 & 0.557 & 0.678 \\ 
   25 & 240 & 217 & 0.904 & 0.542 & 0.678 \\ 
   50 & 215 & 194 & 0.902 & 0.485 & 0.631 \\ 
   100 & 162 & 141 & 0.87 & 0.352 & 0.502 \\  
\end{tabular}
\end{center}
\caption[Cloudbreak accuracy with varying window sizes.]{Accuracy measures for Cloudbreak on the Chromosome 2 simulated data set using different choices of window size. For each window size the same Cloudbreak score threshold was used (1.98).}
\label{chr2AccuracyByWindowSize}
\end{table}
